---
title: "Dietary Patterns and Colorectal Cancer Burden"
output: html_document
date: "2025-11-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#  Title: Dietary Patterns and Colorectal Cancer Burden: Relationship Between Processed Meat Consumption and Cancer Mortality Rates Worldwide

## Team members:
* _Delia Clara Podaru Savu - deliaclara.podaru@estudiants.urv.cat_ 
* _Joao Constantino Muianga - joaoconstantino.muianga@estudiants.urv.cat_ 

## Data Sources:
The data used in this visualization derive from the Colorectal Cancer Global Dataset and Predictions, publicly available on Kaggle. The dataset compiles global country-level statistics on colorectal cancer incidence and mortality alongside demographic and risk-related factors. The dataset is openly accessible and does not require authentication for use in educational or research contexts. Complementary data on country income level were retrieved from the World Bank Open Data API.

## Background:
Colorectal cancer remains one of the leading causes of cancer morbidity and mortality worldwide. The disease burden is influenced by a combination of genetic, demographic, environmental, and lifestyle-related risk factors. Among lifestyle determinants, diet plays a critical role, particularly consumption of processed meats, which has been classified as a Group 1 carcinogen by the International Agency for Research on Cancer (IARC). Understanding the relationship between dietary patterns and cancer outcomes is essential for developing targeted prevention strategies.

## Exploratory Analysis & Data Wrangling:
* The dataset was first inspected for completeness and consistency. Observations lacking values for mortality rate or processed meat consumption were removed (<5% of records). Countries were matched with a World Bank income group classification to contextualize socioeconomic variability. Outliers were examined; extremely small island states with unstable reporting were excluded to prevent distortion.\
*Summary statistics indicated higher colorectal cancer mortality rates in high-income and upper-middle-income countries, regions that also report higher consumption of processed meats. Conversely, many low-income regions showed lower incidence yet increasing mortality over time, likely tied to late detection and limited treatment access. This exploratory pattern led to the decision to focus the explanatory visualization on the relationship between processed meat consumption and mortality, highlighting how lifestyle factors intersect with socioeconomic environments.\
* Aim of the Explanatory Visualization:
The visualization seeks to communicate the association between dietary risk exposure—specifically processed meat consumption—and colorectal cancer mortality rates. The target audience includes policy makers, cancer prevention programs, and public health planners, who influence national dietary guidelines, taxation policies, and awareness campaigns.



```{r cars}
install.packages("pacman")
library(pacman)
pacman::p_load(lubridate, skimr, janitor, rio, here, tidyverse)

colorectal_cancer <- import(here("colorectal_cancer_dataset.csv"))
skim(colorectal_cancer)

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
