---
title: "HDVC Project"
output: html_document
date: "2025-11-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Exploratory Data Analysis

## 1. Load packages and data

```{r}
# load libraries 
#library(tidyverse)
#library(janitor)
#library(scales)
#library(ggthemes)

## we can use pacman, that allow other user to install and load libraries automaticaly
library(pacman)
pacman::p_load(rio,     # importing data
               here,    # relative file pathwyas
               janitor, # data cleaning and tables
               skimr,   # data summary
               scales,  # easily convert proportion to  percent
               corrr,   # correlation analayis for numeric variables
               rstatix, # statistics
               ggExtra, 
               gtsummary, # summary statistic and statistical tests
               epikit,    # age_categories() funtion
               flextable, # converting table to pretty images 
               sf,        # working with map 
               rnaturalearth,
               tidyverse  # data management + ggplot2 graphics
               )


# Import dataset
df_raw <- read.csv("colorectal_cancer_dataset.csv")

# Clean column names 
df <- df_raw %>% clean_names()
```

We used `tidyverse` in order to obtain a a coherent grammar format for an easier data manipulation and plotting.

`clean_names()` is used to standarize te column names and make them more readable like, for example **Cancer_Stage -\> cancer_stage.**

We have kept df_raw (original dataset) and df (working version) separated for better practice.

```{r}
summary(df)

# overview of the dataset
skimr::skim(df)
```

`tidyverse` offers a consistent environment for manipulation and graphics.

`janitor` helps clean column names if necessary.

`glimpse()` and `summary()` are used in order to see and understand the structure of the dataset.

## 2. Data quality check

```{r}
# See the basic structure
glimpse(df)

# Summary statistics for the main variables
df %>%
  select(age, tumor_size_mm, healthcare_costs, incidence_rate_per_100k, mortality_rate_per_100k) %>% summary()

# Check for missing values
colSums(is.na(df))
```

The dataset contains 167,497 patients and 28 variables. All the fields of the dataset are complete, which means we have not detected any missing values (NA). There are numeric variables such as age, tumor size and the healthcare costs, and all of them fall within plausible ranges. This results support the internal consistency and suitability of the data for analysis.

## 3. Variables selected for the project

For this project we focus on variables that describe the patient demographics, disease characteristics and the healthcare context such as age, country, cancer stage, five-year survival, economic classification and healthcare costs.

## 4. Univariate exploration

### 4.1. Age distribution

```{r}
ggplot(df, aes(x= age)) +
  geom_histogram(binwidth = 5, fill = "#80b1d3", color = "white") + geom_vline(xintercept = c(45,85), linetype = "dashed", color = "red") + labs(title = "Age distribution of CC patients", subtitle = "Red dashed lines indicate typical screening age range 45-75 years", x = "Age", y = "Number of patients") + theme_minimal()
```
The age distribution shows that most patients that have had screenings done to them are found in more abundance after the age of 45 years old. 
This pattern is consistent with the known epidemiology of colorectal cancer that defines the following:

- Under 45: Start screenings if the patient presents symptoms.

- Ages 76-85: Screening are a optional but recommended.

- 85+: Screenings are not recommended at this point.

Reference: https://coloncancercoalition.org/screening/when-to-get-screened/ 

### 4.2. Cancer stage distribution

```{r}
df %>% count(cancer_stage) %>% mutate(prop = n / sum(n) * 100)

ggplot(df, aes(x = cancer_stage, fill = cancer_stage)) + geom_bar(color = "black") + scale_fill_manual(values = c("Localized" = "#8ECFC9",
                               "Regional" = "#FFBE7A",
                               "Metastatic" = "#FA7F6F")) + labs(title = "Distribution on CC stages", x = "Cancer Stage", y = "Number of patients") + theme_minimal()
```
Overall, the distribution of the cancer stages is approximately 40% localized, 40% regional and 20% metastatic. This shows that there's a noticeable burden of advanced diasease at diagnosis, even before stratifying by age or country.

### 4.3. Healthcare access distribution

```{r}
df %>% count(healthcare_access) %>% mutate(prop = n/sum(n) *100)
```
The healthcare access is categorised into:
-  ~ 30% of patients reporting high access to healthcare.
- ~ 20% of patients reporting of low access to healthcare.
- ~ 50% of patients reporting moderate to healthcare.

### 4.4. Survival at 5 years (overall)

```{r}
df %>% count(survival_5_years) %>% mutate(prop= n/sum(n)*100)
```
Approximately 60% of the patients survived at least five years after the diagnosis. This will be further stratified by stage and healthcare access in order to better understand the way these factor affect the prognosis.

## 5. Create derived variables 

### 5.1. Age groups

```{r}
# Define age groups
df_new <- df %>% 
  mutate(age_group = cut(age,
                         breaks = c(30,39, 40, 59, 69, 79, 89),labels = c("30-39", "40-49", "50-59", "60-69", "70-79", "80-89" ),include.lowest = TRUE,right= TRUE))
```

```{r}
table(df_new$age_group, useNA = "ifany")
```


Since there are no patients younger than 30, the age groups where defined from 30 years onwards in ten year intervarls. This choice balances granularity with interpretability and aligns with age banding in epidemiological analyses.

### 5.2. Factor ordering 

```{r}
df_new <- df_new %>% mutate(
  cancer_stage = factor(cancer_stage, levels = c("Localized", "Regional", "Metastatic")),
  healthcare_access = factor(healthcare_access, ordered = FALSE),
  country = factor(country)
)
```

The categorical variables from above have been converted to factors and ordered when neccessary like cancer stages from Localized to Metastatic, in order to ensure consistent and clinically meaningful presentation in further plotting.

## 6. Bivariate exploration 

### 6.1. Stage by age group

```{r}
stage_age_table <- df_new %>% group_by(age_group, cancer_stage) %>% summarise(count = n()) %>% mutate(percent = count / sum(count) *100)

print(stage_age_table)
```
JUSTIFICATION

### 6.2. Stage by country

```{r}
stage_by_country_table <- df_new %>% group_by(country, cancer_stage) %>% summarise(count = n()) %>% mutate(percent = count / sum(count) *100)

# Metastatic proportion by country
metastatic_country <- stage_by_country_table %>% filter(cancer_stage == "Metastatic") %>% arrange(desc(percent))

metastatic_country
```
When we examined the distribution of the stages by country, the proportion of metastatic diagnoses where close to 20% in most setting, with a moderst variation in between countries. This results could mean that the late stage diagnosis is not exclusive of only one region, but instead it represents a more global challenge.


### 6.3. Survival by stage and healthcare access

```{r}

```

### Conclusions


# Explanatory Data Analysis



## 1. Summary tables for visualization
```{r}
## Summary table
table_col <- df %>%
  group_by(country, cancer_stage) %>%
  summarise(
    incidence_rate_per_100k  = mean(incidence_rate_per_100k, na.rm = TRUE),
    mortality_rate_per_100k  = mean(mortality_rate_per_100k, na.rm = TRUE),
    cases                    = n(),
    .groups = "drop"
  ) %>%
  # keep one incidence/mortality per country-gender (overall)
  group_by(country) %>%
  mutate(
    inc_overall = mean(incidence_rate_per_100k, na.rm = TRUE),
    mort_overall = mean(mortality_rate_per_100k, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  select(Country = country,  cancer_stage, inc_overall, mort_overall, cases) %>%
  distinct() %>%
  # turn stages into columns of cases
  pivot_wider(
    names_from  = cancer_stage,
    values_from = cases,
    values_fill = 0
  ) %>%
  rename(
    Incidence_per_100k = inc_overall,
    Mortality_per_100k = mort_overall
  ) 

# Flextable
table_col %>%
  flextable() %>%
  autofit() %>%
  theme_booktabs() %>%
  bold(part = "header") %>%
  align(align = "center", part = "all") %>%
  bg(i = seq(1, nrow(table_col), 2), bg = "#f6f6f6") %>%
  color(color = "#132440") %>%
  border_outer() %>%
  border_inner()


```


### 1.1.Summary table - Cancer stage distribution by age group and gender

```{r}
## T-tests
# compare the mean age by stage
df %>%
  select(age, survival_5_years, country, cancer_stage) %>%
  tbl_summary(
    by = cancer_stage,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p})"
    ),  
    missing = "no"
    ) %>%
  add_p() %>%
  modify_header(label = "**Characteristic**") %>%
  modify_caption("**Characteristics by Colorectal Cancer Stage**")


```
### 1.2. Pyramid of aggregate data by age group, gender and cancer stage
```{r}
# aggregate data by age group, gender and cancer stage
df_pyr <- df_new %>%
  group_by(cancer_stage, gender, age_group) %>%
  summarise(n = n(), .groups = "drop") %>%
  # signed counts, one side for each gender
  mutate(n_signed = ifelse(gender== "M", -n, n),
         # make sure age_goupu is ordered corretly
         age_group= factor(age_group, levels = sort(unique(age_group))))

# plot pyramid - split by gender, and facet bu cancer stage
ggplot(df_pyr, aes(x = age_group, y=n_signed,  fill = gender)) +
  geom_col(width = 0.8) +
  coord_flip() +
  scale_y_continuous(labels = abs) +
  facet_wrap( ~ cancer_stage, ncol = 2) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Cancer stage distribution by age group",
    x = "Age group",
    y = "Number of patients",
    fill = "Gender"
  )



```


### 2.3. Cancer stage distribution by country
```{r}
ggplot(df, aes(x = country, fill = cancer_stage)) +
  geom_bar(width = 0.9) +
    scale_fill_manual(values = c("Localized" = "#8ECFC9","Regional" = "#FFBE7A","Metastatic" = "#FA7F6F"))+
  labs(
    title = "Cancer stage distribution by country",
    x = "Country",
    y = "Number of patients",
    fill = "Cancer Stage"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))


```
### 2.3.1 Cancer stage distribution faceted by country
```{r}
# facet by country   
ggplot(df, aes(x = cancer_stage, fill = cancer_stage)) +
  geom_bar() +
  scale_fill_manual(values = c("Localized" = "#8ECFC9","Regional" = "#FFBE7A","Metastatic" = "#FA7F6F"))+
  theme(axis.text.x = element_blank())+
  facet_wrap(~ country) +
  labs(
    title = "Cancer stage distribution by country",
    x = "Cancer stage",
    y = "Number of patients",
    fill = "Cancer Stage"
  ) +
  theme_minimal(base_size = 14)

```


# Survival summary by stage and healthcare access

## 2. Visualizations

### 2.1. Stage distribution by age group
```{r}
ggplot(data = df_new, mapping = aes(y = age, x = cancer_stage, fill = gender)) +
  geom_boxplot(alpha = 0.7) +
  theme(legend.position = "none") +
  theme_minimal(base_size = 14) +
  labs(title = "Stage Distribution by Age",
       x = "Stage",
       y = "Age",
       fill = "Gender")
```


### 2.2. Geographic variation in metastatic diagnosis (as an idea)

```{r}
world <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")

metastatic <- df %>%
  group_by(country) %>%
  summarise(meta_cases = sum(cancer_stage == "Metastatic"), na.rm = T)

map_df <- world %>%
  left_join(metastatic, by = c("name" = "country"))

ggplot(map_df) +
  geom_sf(aes(fill = meta_cases)) +
  scale_fill_gradient(low = "#8ECFC9", high = "#FA7F6F") +
  labs(
    title = "Geographic Variation in Metastatic Diagnosis",
    fill = "% Metastatic"
  ) +
  theme_minimal()
```


### 2.3. Survival by stage and healthcare access

```{r}
df_surv <- df %>%
  mutate(
    surv5 = ifelse(survival_5_years == "Yes", 1, 0),
    cancer_stage = factor(cancer_stage,
                          levels = c("Localized", "Regional", "Metastatic")),
    healthcare_access = factor(healthcare_access,
                               levels = c("Low", "Moderate", "High"))
  )

# descriptive table of survival by stage and healthcare access
# this gives summary of stage, access, 5-year survivors, 5 year survival(%)
tab_stage_access <- df_surv %>%
  group_by(cancer_stage, healthcare_access) %>%
  summarise(
    n = n(),
    survivors_5y = sum(surv5, na.rm = TRUE),
    surv_5y_pct = round(100 * mean(surv5, na.rm = TRUE), 1),
    .groups = "drop"
  )

tab_stage_access %>%
  flextable() %>%
  set_header_labels(
    cancer_stage = "Stage",
    healthcare_access = "Healthcare access",
    n = "N",
    survivors_5y = "5-year survivors",
    surv_5y_pct = "5-year survival (%)"
  ) %>%
  autofit() %>%
  theme_booktabs() %>%
  bold(part = "header") %>%
  align(align = "center", part = "all")

```
_ * The table shows 5-year survival in relation to cancer stage and healthcare access.
* As expected, patients with metastatic disease showed substantially lower 5-year survival compared with localized and regional stages.
* Across all stages, patients with low healthcare access had reduced 5-year survival compared to those with high access, even after adjusting for age and gender in a logistic regression model.
* The interaction between stage and access suggested that the negative impact of poor access is particularly pronounced among patients with advanced (metastatic) disease._


```{r}
## 5 year survival by stage and healthcare access

ggplot(tab_stage_access,
       aes(x = healthcare_access,
           y = surv_5y_pct,
           fill = healthcare_access)) +
  geom_col(width = 0.8) +
  facet_wrap(~ cancer_stage) +
  labs(
    title = "5-year survival by stage and healthcare access",
    x = "Healthcare access",
    y = "5-year survival (%)",
    fill = "Access"
  ) +
  scale_fill_manual(
    values = c(
      "Low" = "#FA7F6F",
      "Moderate" = "#FFBE7A",
      "High" = "#8ECFC9"
    )
  ) +
  theme_minimal(base_size = 14)

```

### Effect of stage & access on 5-year survival
Main effect model
```{r}
model1 <- glm(
  surv5 ~ cancer_stage + healthcare_access + age + gender,
  data = df_surv,
  family = binomial
)

summary(model1)

```
### Effect model with interaction stage X access
```{r}
model2 <- glm(
  surv5 ~ cancer_stage + healthcare_access + age + gender,
  data = df_surv,
  family = binomial()
)

summary(model2)
```

### ORs for survivel by stage and access levels
```{r}
library(broom)
broom::tidy(model2, exponentiate = TRUE, conf.int = TRUE)
```




